---
title: "Data Visualization - Static and Interactive Graphics using R"
author: "Brandon LeBeau"
date: "February 21, 2019"
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: pygments
    transition: fade
    mathjax: null
    includes:
      in_header: custom.css
    reveal_options:
      slideNumber: true
      conrols: false
---

# Workshop Logistics

## About Me
- I'm an Assistant Professor in the College of Education
    + I enjoy model building, particularly longitudinal models, and statistical programming.
- I've used R for over 10 years.
    + I have 4 R packages, 3 on CRAN, 1 on GitHub
        * simglm
        * pdfsearch
        * highlightHTML
        * SPSStoR
- GitHub Repository for this workshop: <https://github.com/lebebr01/iowa_ds_graphics>

## Why teach the tidyverse
- The tidyverse is a series of packages developed by Hadley Wickham and his team at RStudio. <https://www.tidyverse.org/>
- I teach/use the tidyverse for 3 major reasons:
    + Simple functions that do one thing well
    + Consistent implementations across functions within tidyverse (i.e. common APIs)
    + Provides a framework for data manipulation
    
# Static Graphics

## Course Setup
```{r setup_chunks, echo = FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=6, fig.cap = NULL) 
```

```{r setup, message = FALSE, eval = FALSE}
install.packages("tidyverse")
```

```{r library}
library(tidyverse)
```

## Explore Data
```{r data, echo = FALSE, results = 'asis'}
head(data.frame(midwest))
```

## First ggplot
```{r plot1}
ggplot(data = midwest) +
  geom_point(mapping = aes(x = popdensity, y = percollege))
```

## Equivalent Code
```{r plot1_reduced}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege))
```

## Your Turn
1. Try plotting `popdensity` by `state`.
2. Try plotting `county` by `state`. 
    + Does this plot work?
3. Bonus: Try just using the `ggplot(data = midwest)` from above. 
    + What do you get? 
    + Does this make sense?
    
## Add Aesthetics
```{r aesthetic}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state))
```

## Global Aesthetics
```{r global_aes}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege), color = 'pink')
```

## Your Turn
1. Instead of using colors, make the shape of the points different for each state.
2. Instead of color, use `alpha` instead. 
    + What does this do to the plot?
3. Try the following command: `colors()`. 
    + Try a few colors to find your favorite.
4. What happens if you use the following code:

```{r glob_aes, fig.show = 'hide'}
ggplot(midwest) + 
  geom_point(aes(x = popdensity, y = percollege, color = 'green'))
```

## Additional Geoms
```{r smooth, message = FALSE}
ggplot(midwest) +
  geom_smooth(aes(x = popdensity, y = percollege))
```

## Add more Aesthetics
```{r smooth_states, message = FALSE}
ggplot(midwest) +
  geom_smooth(aes(x = popdensity, y = percollege, linetype = state), 
              se = FALSE)
```

## Your Turn
1. It is possible to combine geoms, which we will do next, but try it first. Try to recreate this plot.
```{r combine, echo = FALSE, message = FALSE}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state)) +
  geom_smooth(aes(x = popdensity, y = percollege, color = state), 
              se = FALSE)
```

## Layered ggplot
```{r combine_geoms, message = FALSE}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state)) +
  geom_smooth(aes(x = popdensity, y = percollege, color = state), 
              se = FALSE)
```

## Remove duplicate aesthetics
```{r two_geoms, message = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

## Your Turn
1. Can you recreate the following figure?
```{r differ_aes, message = FALSE, echo = FALSE}
ggplot(midwest, aes(x = popdensity, y = percollege)) +
  geom_point(aes(color = state)) +
  geom_smooth(se = FALSE)
```

## Brief plot customization
```{r breaks_x, eval = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

## Brief plot customization Output
```{r breaks_x2, echo = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

## Change plot theme
```{r theme_bw, message = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() +
  geom_smooth(se = FALSE) + 
  theme_bw()
```

## More themes
+ Themes in ggplot2: <http://ggplot2.tidyverse.org/reference/ggtheme.html>
+ Themes from ggthemes package: <https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html>

## Base plot for reference
```{r base_plot}
p1 <- ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  theme_bw()
```

## Add plot title or subtitle
```{r title_subtitle_ggplot2}
p1 + 
  labs(title = "Percent College Educated by Population Density",
       subtitle = "County level data for five midwest states")
```

## Color Options
```{r grey_color}
p1 + scale_color_grey("State")
```

## Using colorbrewer2.org
+ <http://colorbrewer2.org>

```{r color_brewer}
p1 + scale_color_brewer("State", palette = 'Dark2')
```

## Two additional color options
+ viridis: <https://github.com/sjmgarnier/viridis>
+ scico: <https://github.com/thomasp85/scico>

## viridis colors
```{r viridis, message = FALSE}
library(viridis)
p1 + scale_color_viridis(discrete = TRUE)
```

## viridis colors
```{r viridis2}
p1 + scale_color_viridis(option = 'cividis', discrete = TRUE)
```

## Zoom in on a plot
```{r zoom, eval = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  coord_cartesian(xlim = c(0, 15000))
```

## Zoom in on a plot output
```{r zoom_out, echo = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  coord_cartesian(xlim = c(0, 15000))
```

## Zoom using `scale_x_continuous` - Bad Practice
```{r zoom_x, eval = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density", limits = c(0, 15000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

## Comparing output
```{r zoom_x_output, echo = FALSE, message = FALSE}
library(patchwork)

p1 <- ggplot(data = midwest, 
             aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density", limits = c(0, 15000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")

p2 <- ggplot(data = midwest, 
             aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") +
  coord_cartesian(xlim = c(0, 15000))

p1 + p2
```

## Lord of the Rings Data
- Data from Jenny Bryan: <https://github.com/jennybc/lotr>

```{r read_in_lotr, error = FALSE}
lotr <- read_tsv('https://raw.githubusercontent.com/jennybc/lotr/master/lotr_clean.tsv')
```

## View LOTR
```{r view_lotr}
head(lotr)
```

## Geoms for single variables
```{r histogram}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram() + 
  theme_bw()
```

## Customize histogram
```{r cust_hist}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(bins = 20) + 
  theme_bw()
```

## Customize histogram 2
```{r cust_hist2}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw()
```

## Histograms by other variables - likely not useful
```{r hist_film}
ggplot(lotr, aes(x = Words, color = Film)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw()
```

## Histograms by other variables - one alternative
```{r hist_film_alt}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw() + 
  facet_wrap(~ Film)
```

## Your Turn
1. With more than two groups, histograms are difficult to interpret due to overlap. Instead, use the `geom_density` to create a density plot for `Words` for each film.
2. Using `geom_boxplot`, create boxplots with `Words` as the y variable and `Film` as the x variable. Bonus: facet this plot by the variable `Race`. Bonus2: Zoom in on the bulk of the data.

## Rotation of axis labels
```{r rotate}
ggplot(lotr, aes(x = Film, y = Words)) + 
  geom_boxplot() + 
  facet_wrap(~ Race) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```

## Many times `coord_flip` is better
```{r flip}
ggplot(lotr, aes(x = Film, y = Words)) + 
  geom_boxplot() + 
  facet_wrap(~ Race) + 
  theme_bw() + 
  coord_flip()
```

## Bar graphs
```{r simple_bar}
ggplot(lotr, aes(x = Race)) + 
  geom_bar() + 
  theme_bw()
```

## Add aesthetic
```{r bar_fill}
ggplot(lotr, aes(x = Race)) +
  geom_bar(aes(fill = Film)) + 
  theme_bw()
```

## Stacked Bars Relative
```{r stacked}
ggplot(lotr, aes(x = Race)) + 
  geom_bar(aes(fill = Film), position = 'fill') + 
  theme_bw() + 
  ylab("Proportion")
```

## Dodged Bars
```{r}
ggplot(lotr, aes(x = Race)) + 
  geom_bar(aes(fill = Film), position = 'dodge') + 
  theme_bw()
```

## Change Bar Col bar_coloror
```{r bar_color}
ggplot(lotr, aes(x = Race)) + 
  geom_bar(aes(fill = Film), position = 'fill') + 
  theme_bw() + 
  ylab("Proportion") + 
  scale_fill_viridis(option = 'cividis', discrete = TRUE)
```

## Your Turn
1. Using the gss_cat data, create a bar chart of the variable `partyid`.
2. Add the variable `marital` to the bar chart created in step 1. Do you prefer a stacked or dodged version?
3. Take steps to make one of the plots above close to publication quality.

## Additional ggplot2 resources
+ ggplot2 website: <http://docs.ggplot2.org/current/index.html>
+ ggplot2 book: <http://www.springer.com/us/book/9780387981413>
+ R graphics cookbook: <http://www.cookbook-r.com/Graphs/>

## Additional R Resources
+ R for Data Science: <http://r4ds.had.co.nz/>

# Interactive Graphics

## Why Interactive Graphics
* Why interactive graphics?
    + Created specifically for the web.
    + Can focus, explore, zoom, or remove data at will.
    + Allows users to customize their experience.
    + It is fun!

## Interactive graphics with plotly
```{r install, eval = FALSE}
install.packages("plotly")
```

## First Interactive Plot
```{r first_plotly, message = FALSE, results = 'asis', fig.height = 6}
library(plotly)
p <- ggplot(data = midwest) +
  geom_point(mapping = aes(x = popdensity, y = percollege))
ggplotly(p, width = 900, height = 500)
```

## Customized Plot
```{r custom_plotly, warning = FALSE, fig.height = 6}
p <- ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  theme_bw()
```

## Interactive Output
```{r int_out, results = 'asis', fig.height = 6}
ggplotly(p, width = 900)
```


## Your Turn
1. Using the `starwars` data, create a static ggplot and use the `ggplotly` function to turn it interactive. 

## Create plotly by hand
```{r plotly_by_hand, results = 'asis', fig.height = 6}
plot_ly(lotr, x = ~Words, width = 900) %>% add_histogram()
```

## Subplots Code
```{r subplots, results = 'asis'}
one_plot <- function(d) {
  plot_ly(d, x = ~Words, width = 900) %>%
    add_histogram() %>%
    add_annotations(
      ~unique(Film), x = 0.5, y = 1, 
      xref = "paper", yref = "paper", showarrow = FALSE
    )
}

plotly_out <- lotr %>%
  split(.$Film) %>%
  lapply(one_plot) %>% 
  subplot(nrows = 1, shareX = TRUE, titleX = FALSE) %>%
  hide_legend()
```

## Subplots Output
```{r sub_out, results = 'asis', fig.height = 6}
plotly_out
```


## Grouped bar plot
```{r plotly_group, results = 'asis', fig.height = 6}
plot_ly(lotr, x = ~Race, color = ~Film, width = 900) %>% add_histogram()
```

## Plot of proportions code
```{r plotly_proportions, message = FALSE, results = 'asis'}
## number of diamonds by cut and clarity (n)
lotr_count <- count(lotr, Race, Film)
## number of diamonds by cut (nn)
lotr_prop <- left_join(lotr_count, count(lotr_count, Race, wt = n), 
                       by = 'Race')

plotly_out <-lotr_prop %>%
  mutate(prop = n.x / n.y) %>%
  plot_ly(x = ~Race, y = ~prop, color = ~Film, width = 900) %>%
  add_bars() %>%
  layout(barmode = "stack")
```

## Plot of proportions output
```{r prop_code, results = 'asis', fig.height = 6}
plotly_out
```


## Your Turn
1. Using the `gss_cat` data, create a histrogram for the `tvhours` variable. 
2. Using the `gss_cat` data, create a bar chart showing the `partyid` variable by the `marital` status.

## Scatterplots by Hand
```{r plotly_scatter, warning = FALSE, results = 'asis', fig.height = 6}
plot_ly(midwest, x = ~popdensity, y = ~percollege, width = 900) %>%
  add_markers()
```

## Change symbol
```{r plotly_symbol, results = 'asis', fig.height = 6}
plot_ly(midwest, x = ~popdensity, y = ~percollege, width = 900) %>%
  add_markers(symbol = ~state)
```

## Change color
```{r plotly_color, results = 'asis', fig.height = 6}
plot_ly(midwest, x = ~popdensity, y = ~percollege, width = 900) %>%
  add_markers(color = ~state, colors = viridis::viridis(5))
```

## Line Graph
```{r plotly_line, results = 'asis', fig.height = 6}
storms_yearly <- storms %>%
  group_by(year) %>%
  summarise(num = length(unique(name)))

plot_ly(storms_yearly, x = ~year, y = ~num, width = 900, height = 500) %>%
  add_lines()
```

## Your Turn
1. Using the `gss_cat` data, create a scatterplot showing the `age` and `tvhours` variables.
2. Compute the average time spent watching tv by year and marital status. Then, plot the average time spent watching tv by year and marital status.

## Highcharter; Highcharts for R
```{r install_highcharter, eval = FALSE}
devtools::install_github("jbkunst/highcharter")
```

## Load `highcharter`
```{r hchart1}
library(highcharter)

lotr_count <- lotr %>%
  count(Film, Race)
head(lotr_count)
```

## `hchart` function
```{r hchart1a, results = 'asis'}
hchart(lotr_count, "column", hcaes(x = Race, y = n, group = Film))
```

## A second `hchart`
```{r hchart2, results = 'asis'}
hchart(midwest, "scatter", hcaes(x = popdensity, y = percollege, group = state))
```

## Histogram
```{r hchart_hist, results = 'asis'}
hchart(lotr$Words)
```

## Your Turn
1. Using the `hchart` function, create a bar chart or histogram with the `gss_cat` data.
2. Using the `hchart` function, create a scatterplot with the `gss_cat` data.

## Build Highcharts from scratch
```{r hc_scratch}
hc <- highchart() %>%
  hc_xAxis(categories = lotr_count$Race) %>%
  hc_add_series(name = 'The Fellowship Of The Ring', 
                data = filter(lotr_count, Film == 'The Fellowship Of The Ring')$n) %>% 
  hc_add_series(name = 'The Two Towers', 
                data = filter(lotr_count, Film == 'The Two Towers')$n) %>%
  hc_add_series(name = 'The Return Of The King', 
                data = filter(lotr_count, Film == 'The Return Of The King')$n)
```

## Build highcharts from scratch output
```{r hc_s, reuslts = 'asis'}
hc
```


## Change Chart type
```{r hc_chart, results = 'asis'}
hc <- hc %>%
  hc_chart(type = 'column')
hc
```

## Change Colors
```{r hc_change_colors, results = 'asis'}
hc <- hc %>%
  hc_colors(substr(viridis(3), 0, 7))
hc
```

## Modify Axes
```{r hc_axis, results = 'asis'}
hc <- hc %>%
  hc_xAxis(title = list(text = "Race")) %>%
  hc_yAxis(title = list(text = "Number of Words Spoken"),
           showLastLabel = FALSE)
hc
```

## Add title, subtitle, move legend
```{r hc_modify}
hc <- hc %>%
  hc_title(text = 'Number of Words Spoken in Lord of the Rings Films',
           align = 'left') %>%
  hc_subtitle(text = 'Broken down by <i>Film</i> and <b>Race</b>', 
              align = 'left') %>%
  hc_legend(align = 'right', verticalAlign = 'top', layout = 'vertical',
            x = 0, y = 80) %>%
  hc_exporting(enabled = TRUE)
```

## Add title, subtitle, move legend output
```{r hc_mod2, results = 'asis'}
hc
```


## Your Turn
1. Build up a plot from scratch, getting the figure close to publication quality using the `gss_cat` data.

## Correlation Matrices
```{r correlation, results = 'asis'}
select(storms, wind, pressure, ts_diameter, hu_diameter) %>%
  cor(use = "pairwise.complete.obs") %>%
  hchart()
```

## Leaflet Example
```{r leaflet, results = 'asis'}
library(leaflet)

storms %>%
  filter(name %in% c('Ike', 'Katrina'), year > 2000) %>%
  leaflet() %>%
  addTiles() %>%
  addCircles(lng = ~long, lat = ~lat, popup = ~name, weight = 1,
             radius = ~wind*1000)
```

## Additional Resources
* plotly for R book: <https://plotly-book.cpsievert.me/>
* plotly: <https://plot.ly/>
* highcharter: <http://jkunst.com/highcharter/index.html>
* highcharts: <https://www.highcharts.com/>
* htmlwidgets: <https://www.htmlwidgets.org/>

