---
title: "Data Visualization - Static and Interactive using R"
author: "Brandon LeBeau"
date: "June 5, 2018"
output: html_notebook
---

# Course Setup
```{r setup_chunks, echo = FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=6, fig.cap = NULL) 
```

```{r setup, message = FALSE, eval = FALSE}
install.packages("tidyverse")
```

```{r library}
library(tidyverse)
```

# Explore Data
```{r data, echo = FALSE, results = 'asis'}
midwest
```

# First ggplot
```{r plot1}
ggplot(data = midwest) +
  geom_point(mapping = aes(x = popdensity, y = percollege))
```

# Equivalent Code
```{r plot1_reduced}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege))
```

# Your Turn
1. Try plotting `popdensity` by `state`.
2. Try plotting `county` by `state`. 
    + Does this plot work?
3. Bonus: Try just using the `ggplot(data = midwest)` from above. 
    + What do you get? 
    + Does this make sense?

# Add Aesthetics
```{r aesthetic}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state))
```

# Global Aesthetics
```{r global_aes}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege), color = 'pink')
```

# Your Turn
1. Instead of using colors, make the shape of the points different for each state.
2. Instead of color, use `alpha` instead. 
    + What does this do to the plot?
3. Try the following command: `colors()`. 
    + Try a few colors to find your favorite.
4. What happens if you use the following code:
```{r glob_aes, fig.show = 'hide'}
ggplot(midwest) + 
  geom_point(aes(x = popdensity, y = percollege, color = 'green'))
```

# Additional Geoms
```{r smooth, message = FALSE}
ggplot(midwest) +
  geom_smooth(aes(x = popdensity, y = percollege))
```

# Add more Aesthetics
```{r smooth_states, message = FALSE}
ggplot(midwest) +
  geom_smooth(aes(x = popdensity, y = percollege, linetype = state), 
              se = FALSE)
```

# Your Turn
1. It is possible to combine geoms, which we will do next, but try it first. Try to recreate this plot.
```{r combine, echo = FALSE, message = FALSE}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state)) +
  geom_smooth(aes(x = popdensity, y = percollege, color = state), 
              se = FALSE)
```

# Layered ggplot
```{r combine_geoms, message = FALSE}
ggplot(midwest) +
  geom_point(aes(x = popdensity, y = percollege, color = state)) +
  geom_smooth(aes(x = popdensity, y = percollege, color = state), 
              se = FALSE)
```

# Remove duplicate aesthetics
```{r two_geoms, message = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

# Your Turn
1. Can you recreate the following figure?
```{r differ_aes, message = FALSE, echo = FALSE}
ggplot(midwest, aes(x = popdensity, y = percollege)) +
  geom_point(aes(color = state)) +
  geom_smooth(se = FALSE)
```

# Brief plot customization
```{r breaks_x, eval = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

# Brief plot customization Output
```{r breaks_x2, echo = FALSE}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

# Change plot theme
```{r theme_bw}
ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() +
  geom_smooth(se = FALSE) + 
  theme_bw()
```

# More themes
+ Themes in ggplot2: <http://ggplot2.tidyverse.org/reference/ggtheme.html>
+ Themes from ggthemes package: <https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html>

# Base plot for reference
```{r base_plot}
p1 <- ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  theme_bw()
```

# Color Options
```{r grey_color}
p1 + scale_color_grey("State")
```

# Using colorbrewer2.org
+ <http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3>

```{r color_brewer}
p1 + scale_color_brewer("State", palette = 'Dark2')
```

# Two additional color options
+ viridis: <https://github.com/sjmgarnier/viridis>
+ scico: <https://github.com/thomasp85/scico>

# viridis colors
```{r viridis}
library(viridis)
p1 + scale_color_viridis(discrete = TRUE)
```

# viridis colors
```{r viridis2}
p1 + scale_color_viridis(option = 'cividis', discrete = TRUE)
```

# Zoom in on a plot
```{r zoom, eval = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  coord_cartesian(xlim = c(0, 15000))
```

# Zoom in on a plot output
```{r zoom_out, echo = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  coord_cartesian(xlim = c(0, 15000))
```

# Zoom using `scale_x_continuous` - Bad Practice
```{r zoom_x, eval = FALSE}
ggplot(data = midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density", limits = c(0, 15000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")
```

# Comparing output
```{r zoom_x_output, echo = FALSE}
library(patchwork)

p1 <- ggplot(data = midwest, 
             aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density", limits = c(0, 15000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State")

p2 <- ggplot(data = midwest, 
             aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  scale_x_continuous("Population Density") + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") +
  coord_cartesian(xlim = c(0, 15000))

p1 + p2
```

# Lord of the Rings Data
- Data from Jenny Bryan: <https://github.com/jennybc/lotr>

```{r read_in_lotr}
lotr <- read_tsv('https://raw.githubusercontent.com/jennybc/lotr/master/lotr_clean.tsv')
lotr
```

# Geoms for single variables
```{r histogram}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram() + 
  theme_bw()
```

# Customize histogram
```{r cust_hist}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(bins = 20) + 
  theme_bw()
```

# Customize histogram 2
```{r cust_hist2}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw()
```

# Histograms by other variables - likely not useful
```{r hist_film}
ggplot(lotr, aes(x = Words, color = Film)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw()
```

# Histograms by other variables - one alternative
```{r hist_film_alt}
ggplot(lotr, aes(x = Words)) + 
  geom_histogram(binwidth = 25) + 
  theme_bw() + 
  facet_wrap(~ Film)
```

# Your Turn
1. With more than two groups, histograms are difficult to interpret due to overlap. Instead, use the `geom_density` to create a density plot for `Words` for each film.

2. Using `geom_boxplot`, create boxplots with `Words` as the y variable and `Film` as the x variable. Bonus: facet this plot by the variable `Race`. Bonus2: Zoom in on the bulk of the data.

# Rotation of axis labels
```{r rotate}
ggplot(lotr, aes(x = Film, y = Words)) + 
  geom_boxplot() + 
  facet_wrap(~ Race) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))
```

# Many times `coord_flip` is better
```{r flip}
ggplot(lotr, aes(x = Film, y = Words)) + 
  geom_boxplot() + 
  facet_wrap(~ Race) + 
  theme_bw() + 
  coord_flip()
```

# Bar graphs
```{r simple_bar}
ggplot(lotr, aes(x = Race)) + 
  geom_bar() + 
  theme_bw()
```

# Add aesthetic
```{r bar_fill}
ggplot(lotr, aes(x = Race)) +
  geom_bar(aes(fill = Film)) + 
  theme_bw()
```

# Stacked Bars
```{r stacked}
ggplot(lotr, aes(x = Race)) + 
  geom_bar(aes(fill = Film), position = 'fill') + 
  theme_bw()
```

# Dodged Bars
```{r}
ggplot(lotr, aes(x = Race)) + 
  geom_bar(aes(fill = Film), position = 'dodge') + 
  theme_bw()
```

# Your Turn
1. Using the gss_cat data, create a bar chart of the variable `partyid`.
2. Add the variable `marital` to the bar chart created in step 1. Do you prefer a stacked or dodged version?
3. Take steps to make one of the plots above close to publication quality.


# Additional ggplot2 resources
+ ggplot2 website: <http://docs.ggplot2.org/current/index.html>
+ ggplot2 book: <http://www.springer.com/us/book/9780387981413>
+ R graphics cookbook: <http://www.cookbook-r.com/Graphs/>

# Interactive graphics
```{r install, eval = FALSE}
install.packages("plotly")
```

# First Interactive Plot
```{r first_plotly}
p <- ggplot(data = midwest) +
  geom_point(mapping = aes(x = popdensity, y = percollege))
ggplotly(p)
```

# Customized Interactive Plot
```{r custom_plotly, warning = FALSE}
p <- ggplot(midwest, 
       aes(x = popdensity, y = percollege, color = state)) +
  geom_point() + 
  scale_x_continuous("Population Density", 
                     breaks = seq(0, 80000, 20000)) + 
  scale_y_continuous("Percent College Graduates") + 
  scale_color_discrete("State") + 
  theme_bw()
ggplotly(p)
```

# Create plotly by hand
```{r plotly_by_hand}
plot_ly(lotr, x = ~Words) %>% add_histogram()
```

# Subplots
```{r subplots}
one_plot <- function(d) {
  plot_ly(d, x = ~Words) %>%
    add_histogram() %>%
    add_annotations(
      ~unique(Film), x = 0.5, y = 1, 
      xref = "paper", yref = "paper", showarrow = FALSE
    )
}

lotr %>%
  split(.$Film) %>%
  lapply(one_plot) %>% 
  subplot(nrows = 1, shareX = TRUE, titleX = FALSE) %>%
  hide_legend()
```


# Grouped bar plot
```{r plotly_group}
plot_ly(lotr, x = ~Race, color = ~Film) %>% add_histogram()
```

# Plot of proportions
```{r plotly_proportions}
# number of diamonds by cut and clarity (n)
lotr_count <- count(lotr, Race, Film)
# number of diamonds by cut (nn)
lotr_prop <- left_join(lotr_count, count(lotr_count, Race, wt = n))
lotr_prop %>%
  mutate(prop = n / nn) %>%
  plot_ly(x = ~Race, y = ~prop, color = ~Film) %>%
  add_bars() %>%
  layout(barmode = "stack")
```







# Additional Interactive Graphics Resources
* r2d3 package: <https://rstudio.github.io/r2d3/>
* plotly for R book: <https://plotly-book.cpsievert.me/>
* htmlwidgets: <https://www.htmlwidgets.org/>
* htmlwidgets showcase: <https://www.htmlwidgets.org/showcase_leaflet.html>


# Additional Resources
+ R for Data Science: <http://r4ds.had.co.nz/>